---
title: "eds-223-hw2"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

# Exploring Patterns of Environmental Justic in Los Angeles County
Primary Objective
Los Angeles County is the most populous county in the United States as well as the state of California, with about 9.8 million residents as of 2023 (https://datausa.io/profile/geo/los-angeles-county-ca).

## Part 1: Legacy of redlining in current environmental (in)justice

Your first task is to explore historical redlining in Los Angeles and
its legacy on present-day environmental justice.

Description 
For this assignment, you must complete the following:

1. Create a map of historical redlining neighborhoods, including:
- neighborhoods colored by HOLC grade an appropriate base map 
- a clear title and legend 

```{r}
#| include: false
# Import libraries
library(sf)
library(here)
library(tidyverse)
library(ggplot2)
library(tmap)
library(kableExtra)

```


```{r}
#| output: false
# Now, load in the data
ej_screen <- st_read(here::here("data","ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
birds <- st_read(here::here("data","gbif-birds-LA", "gbif-birds-LA.shp"))
inequal <- st_read(here::here("data","mapping-inequality", "mapping-inequality-los-angeles.json"))
```


```{r}
# Check CRS for all
st_crs(ej_screen) # WGS 84 / Pseudo-Mercator 
st_crs(birds)  # WGS 84 
st_crs(inequal) # WGS 84 

# Transform to match same CRS as ej_screen
birds <- st_transform(birds, st_crs(ej_screen))
inequal <- st_transform(inequal, st_crs(ej_screen))
```



```{r}
# Create a map of historical redlining neighborhoods, including:
# - neighborhoods colored by HOLC grade 
# - an appropriate base map 
# - a clear title and legend 
map1 <- tm_shape(inequal) +
  tm_tiles("CartoDB.Positron") + #Establish a basemap with tm_tiles using a view from https://leaflet-extras.github.io/leaflet-providers/preview/
  tm_polygons(
    col = "grade",
    palette = "brewer.Reds",
    alpha = 0.7,                      
    border.col = "black",
    title = "HOLC Grade")   +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_graticules() +
  tm_layout(
    main.title = "Historical Redlining in Los Angeles (HOLC)", 
    title.size = 1,
    #outer.margins = c(0.1, 0.05, 0.05, 0.05),
    legend.outside = TRUE,
    legend.outside.position = "right",
    component.autoscale = FALSE
  )

map1

tmap_save(map1, "figs/LA_HOC.png")

```


2. Create a table summarizing: 
- the percentage of census block groups that fall within each HOLC grade 
- Also include the percent of census block groups that don’t fall within a HOLC grade 
- Hint: The HOLC data contains the grades and the EJScreen data contains the census
blocks, so you will need to combine the data spatially before doing
summary statistics. Once you combine and no longer need the geometries,
you can use st_drop_geometry(). 

```{r}
#  Hint: The HOLC data contains the grades and the EJScreen data contains the census
# blocks, so you will need to combine the data spatially before doing
# summary statistics. Once you combine and no longer need the geometries,
# you can use st_drop_geometry(). 
str(ej_screen)
str(inequal)

# thought process: 
# first, need to combine data spatially via spatial join (left to keep observations in x + observations in y that are found in x), then drop geoms...  assign HOLC grade to each census block

# Filter for 
ej_la <- ej_screen %>% filter(STATE_NAME == "California", CNTY_NAME == "Los Angeles County") # Keep all LA census blocks and not just ones that intersect HOLC polygons

census_holc <- st_join(ej_la, inequal, left = T)
census_holc_df <- st_drop_geometry(census_holc)

census_summary <- census_holc_df %>%
  mutate(grade = ifelse(is.na(grade), "No HOLC grade", grade)) %>% 
  group_by(grade) %>% summarise(count = n())%>% # count # of rows in each grade group
  mutate(percent = round(100 * count / sum(count), 2)) # create a new variable to calculate the percent of how much each grade over the total observations represents

census_summary %>% 
  kbl() %>%
  kable_styling() # https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html 


```




3. Create at least two visualizations summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables (you may combine
variables or create separate plots): 

- % low income 
- percentile for Particulate Matter 2.5 
- percentile for low life expectancy 
-Use ggplot for your visualizations! You will first need to calculate mean of each variable grouped by HOLC grade. 

```{r}
holc_summary <- census_holc_df %>%
  group_by(grade) %>% 
  mutate(grade = ifelse(is.na(grade), "No HOLC grade", grade)) %>%
  filter(grade != "No HOLC grade") %>%
  summarise(
    mean_low_income = mean(P_LOWINCPCT, na.rm = TRUE),
    mean_pm25 = mean(P_PM25, na.rm = TRUE),
    mean_low_life_exp = mean(P_LIFEEXPPCT, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "socio_economic",
    values_to = "mean_value"
  ) 

# bar plot: mean value for each variable by HOLC grade
holc_summary %>% ggplot(aes(x = grade, y = mean_value, fill = grade)) +
  geom_col() +
  facet_wrap(~ socio_economic, scales = "free_y") +
  labs(
    x = "HOLC Grade",
    y = "Mean Value",
    title = "Mean EJScreen Indicators by HOLC Grade"
  ) +
  theme_bw() + 
  theme(legend.position = "none") 

# 
ggplot(holc_summary, aes(x = grade, y = mean_value, color = grade)) +
  geom_point(size = 3) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    x = "HOLC Grade",
    y = "Mean Value",
    title = "Mean EJScreen Indicators by HOLC Grade"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


Write a brief paragraph reflecting on these results. Interpret the patterns you observe in your results. Discuss potential relationships between historical redlining grades and current
environmental/socioeconomic conditions. 




## Part 2: Legacy of redlining in biodiversity observations

Description
For this assignment, you must produce the following based on observations from 2021-2023:

1. A figure summarizing the percent of bird observations within redlined neighborhoods within each HOLC grade

Create a visualizations that shows:
- The percentage of bird observations within each HOLC grade
```{r}
# Ensure in same CRS! 
st_crs(birds) == st_crs(inequal) # Yes they match 


# Perform spatial join to assign each bird observation to a HOLC grade polygon
# st_intersects so points on borders are included in spatial 
birds_holc <- st_join(birds, inequal, join = st_intersects, left = FALSE)

# Summarize % of bird observations by HOLC grade
bird_summary <- birds_holc %>%
  st_drop_geometry() %>% #
  group_by(grade) %>%
  summarise(count = n()) %>%
  mutate(percent = round(100 * count / sum(count), 2),
         grade = factor(grade, levels = c("A","B","C","D"))) %>%
  arrange(grade)

print(bird_summary)

# creat a pallete for each grade (percent of bird observations per HOLC grade)
grade_colors <- c("A" = "#2ca25f",
                  "B" = "#99d8c9",
                  "C" = "#fec44f",
                  "D" = "#de2d26")

bird_summary %>% ggplot(aes(x = grade, y = percent, fill = grade)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(percent, "%")), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = grade_colors) +
  labs(
    title = "Percent of Bird Observations (2021–2023) within HOLC Grades",
    x = "HOLC Grade",
    y = "Percent of Bird Observations",
    caption = "Bird observations within Los Angeles HOLC neighborhoods (2021–2023)."
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))



```


- Include an appropiate title, axis labels, and legend
- Hints: Ensure the bird observations and HOLC dataset have matching CRS’, then perform a spatial join to assign each bird observations to a corresponding HOLC grade.
- Spolier alert!! Our results don’t match the findings from Ellis-Soto et al. 2023! Read the abstract of the study. Why might we have obtained different results in our analysis? What did the paper consider that we did not?








