---
title: "eds-223-hw2"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## Part 1: Legacy of redlining in current environmental (in)justice

Your first task is to explore historical redlining in Los Angeles and
its legacy on present-day environmental justice.

Description 
For this assignment, you must complete the following:

1. Create a map of historical redlining neighborhoods, including:
- neighborhoods colored by HOLC grade an appropriate base map 
- a clear title and legend 

```{r}
#| include: false
# Import libraries
library(sf)
library(here)
library(tidyverse)
library(ggplot2)
library(tmap)
```


```{r}
#| include: false
# Now, load in the data
ej_screen <- st_read(here::here("data","ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
birds <- st_read(here::here("data","gbif-birds-LA", "gbif-birds-LA.shp"))
inequal <- st_read(here::here("data","mapping-inequality", "mapping-inequality-los-angeles.json"))
```


```{r}
# Check crs for all
st_crs(ej_screen) # WGS 84 / Pseudo-Mercator 
st_crs(birds)  # WGS 84 
st_crs(inequal) # WGS 84 

# Transform to match same CRS as ej_screen
birds <- st_transform(birds, st_crs(ej_screen))
inequal <- st_transform(inequal, st_crs(ej_screen))
```



```{r}
# Create a map of historical redlining neighborhoods, including:
# - neighborhoods colored by HOLC grade 
# - an appropriate base map 
# - a clear title and legend 
map1 <- tm_shape(inequal) +
  tm_tiles("CartoDB.Positron") + #https://leaflet-extras.github.io/leaflet-providers/preview/
  tm_polygons(
    col = "grade",
    palette = "Reds",
    alpha = 0.7,                      
    border.col = "black",
    title = "HOLC Grade")   +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_graticules() +
  tm_layout(
    main.title = "Historical Redlining in Los Angeles (HOLC)", 
    title.size = 1,
    #outer.margins = c(0.1, 0.05, 0.05, 0.05),
    legend.outside = TRUE,
    legend.outside.position = "right",
    component.autoscale = FALSE
  )

map1

tmap_save(map1, "figs/LA_HOC.png")

```


2. Create a table summarizing: 
- the percentage of census block groups that fall within each HOLC grade 
- Also include the percent of census black groups that don’t fall within a HOLC grade 
- Hint: The HOLC data contains the grades and the EJScreen data contains the census
blocks, so you will need to combine the data spatially before doing
summary statistics. Once you combine and no longer need the geometries,
you can use st_drop_geometry(). 

```{r}
#  Hint: The HOLC data contains the grades and the EJScreen data contains the census
# blocks, so you will need to combine the data spatially before doing
# summary statistics. Once you combine and no longer need the geometries,
# you can use st_drop_geometry(). 





```




3. Create at least two visualizations summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables (you may combine
variables or create separate plots): 

- % low income percentile for Particulate Matter 2.5 
- percentile for low life expectancy Use ggplot for your visualizations! You will first need to calculate mean of each variable grouped by HOLC grade. 

Write a brief paragraph reflecting on these results. Interpret the patterns you observe in your results. Discuss potential relationships between historical redlining grades and current
environmental/socioeconomic conditions. 


## Part 2: Legacy of redlining in biodiversity observations

Description
For this assignment, you must produce the following based on observations from 2021-2023:

1. A figure summarizing the percent of observations within redlined neighborhoods within each HOLC grade

Create a visualizations that shows:
- The percentage of bird observations within each HOLC grade
- Include an appropiate title, axis labels, and legend
- Hints: Ensure the bird observations and HOLC dataset have matching CRS’, then perform a spatial join to assign each bird observations to a corresponding HOLC grade.
- Spolier alert!! Our results don’t match the findings from Ellis-Soto et al. 2023! Read the abstract of the study. Why might we have obtained different results in our analysis? What did the paper consider that we did not?








